---
title: "HW3"
author: "wd2311"
date: "2025-10-14"
output: github_document
---
# Loading libraries
```{r}
library(tidyverse)
library(janitor)
library(p8105.datasets)
library(knitr)
library(forcats)
library(patchwork)
library(scales)
```


# Problem 1
## Set up and brief description
The instacart dataset contains 1,384,617 observations and 15 variables, each 
representing a single product added to a user’s grocery order on the Instacart 
online shopping platform. Each row corresponds to one product within one order, 
identified by variables such as:
order_id (unique order identifier),
product_id and product_name (the specific item purchased),
add_to_cart_order (the position in which the item was added),
reordered (whether the product had been purchased before by the same user),
order_dow and order_hour_of_day (day of week and time the order was placed),
aisle and department (product category classification).
```{r}
data("instacart") 
glimpse(instacart)
n_obs  = nrow(instacart)
n_vars = ncol(instacart)
n_aisles = instacart |> 
  distinct(aisle_id, aisle) |> 
  nrow()

c(n_obs, n_vars, n_aisles)
```

## Description on aisles
There are 134 aisles. The aisles with the greatest number of items ordered are 
“fresh vegetables” (150609 orders) and “fresh fruits” (150473 orders).
```{r}
aisle_counts = 
  instacart |>
  count(aisle, sort = TRUE)

n_aisles = nrow(aisle_counts)
n_aisles

# top aisles
aisle_counts |> 
  slice_head(n = 10) |> 
  kable()
```

## Plot on aisles with more than 10000 items
```{r}
aisle_counts |>
  filter(n > 10000) |>
  mutate(aisle = fct_reorder(aisle, n)) |>
  ggplot(aes(x = n, y = aisle)) +
  geom_col() +
  labs(
    x = "Number of items ordered",
    y = NULL,
    title = "Items ordered by aisle (aisles with > 10,000 items)"
  ) +
  theme_minimal(base_size = 12)
```

## Table for top3 items by aisle
```{r}
top3_by_aisle = 
  instacart |>
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  count(aisle, product_name, sort = TRUE) |>
  group_by(aisle) |>
  slice_max(n, n = 3, with_ties = FALSE) |>
  arrange(aisle, desc(n)) |>
  ungroup()

kable(top3_by_aisle, col.names = c("Aisle", "Product", "Times ordered"))
```

## Table for mean hours of the day
```{r}
dow_labs = c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")

mean_hour_tbl = 
  instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  mutate(dow = factor(order_dow, levels = 0:6, labels = dow_labs)) |>
  group_by(product_name, dow) |>
  summarise(mean_hour = mean(order_hour_of_day), .groups = "drop") |>
  mutate(mean_hour = round(mean_hour, 2)) |>
  pivot_wider(names_from = dow, values_from = mean_hour) |>
  arrange(product_name)

kable(mean_hour_tbl, col.names = c("Product","Sun","Mon","Tue","Wed","Thu","Fri","Sat"))

```


# Problem 2
## Data Set-up
```{r}
zip_meta = 
  read_csv("data/Zip Codes.csv") |> 
  clean_names() 

zori_raw = 
  read_csv("data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv", show_col_types = FALSE) |> 
  clean_names() 

# Check column names
names(zip_meta)
names(zori_raw)
```

## Data preparation 
```{r}
zip_meta = 
  zip_meta |>
  mutate(
    zip = str_pad(as.character(zip_code), 5, pad = "0"),
    county = str_to_title(county) |> str_trim()
  )

county_to_borough <- c(
  "Bronx"          = "Bronx",
  "Kings"          = "Brooklyn",
  "New York"       = "Manhattan",
  "Queens"         = "Queens",
  "Richmond"       = "Staten Island"
)

zip_meta = 
  zip_meta |>
  mutate(borough = recode(county, !!!county_to_borough)) |>
  filter(!is.na(borough)) |>
  select(zip, borough, county, neighborhood)
```

## Turning to long form
```{r}
zori_long = 
  zori_raw |>
  mutate(
    zip = str_pad(as.character(region_name), 5, pad = "0")
  ) |>
  pivot_longer(
    cols = matches("^x\\d{4}_\\d{2}_\\d{2}$"),
    names_to = "date_raw",
    values_to = "zori"
  ) |>
  mutate(
    date = ymd(str_replace_all(str_remove(date_raw, "^x"), "_", "-")),
    year = year(date),
    month = month(date)
  ) |>
  filter(!is.na(date))

nyc = 
  zori_long |>
  left_join(zip_meta |> select(zip, borough), by = "zip") |>
  filter(!is.na(borough))
```

## Monthly Observation
```{r}
nyc_target = 
  nyc |>
  filter(date >= ymd("2015-01-01"), date <= ymd("2024-08-31"))

months_per_zip = 
  nyc_target |>
  count(zip, name = "n_months")

n_116  = months_per_zip |> 
  filter(n_months == 116) |> 
  nrow()
n_lt10 = months_per_zip |> 
  filter(n_months < 10)   |> 
  nrow()

cat("ZIP codes observed exactly 116 times:", n_116, "\n")
cat("ZIP codes observed fewer than 10 times:", n_lt10, "\n")

```

## Table on Borough × Year 
```{r}
dir.create("results", showWarnings = FALSE)

borough_year_tbl = 
  nyc_target |>
  group_by(borough, year) |>
  summarize(avg_rent = mean(zori, na.rm = TRUE), .groups = "drop") |>
  arrange(borough, year)

borough_year_wide = 
  borough_year_tbl |>
  pivot_wider(names_from = year, values_from = avg_rent) |>
  mutate(across(-borough, ~ round(.x, 0))) 

kable(
  borough_year_wide |>
    mutate(across(-borough, ~ dollar(.x))),
  caption = "Average Rental Price by Borough and Year (ZORI)"
)


```


## Plot for all available years
```{r}
nyc_clean = 
  nyc_target |>
  filter(!is.na(zori), is.finite(zori))  

all_years_plot = 
  nyc_clean |>
  group_by(borough, date) |>
  mutate(borough_median = median(zori, na.rm = TRUE)) |>
  ungroup() |>
  ggplot(aes(x = date, y = zori, group = zip)) +
  geom_line(alpha = 0.08) +
  stat_summary(
    aes(group = borough),
    fun = median, geom = "line", linewidth = 1.1
  ) +
  facet_wrap(~ borough, ncol = 2, scales = "free_y") +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "NYC Rental Prices within ZIP Codes (All Years)",
    subtitle = "Thin lines: ZIP series · Thick line: Borough median ZORI",
    x = NULL, y = "ZORI ($)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())

ggsave("results/plot_all_years_by_borough.png", all_years_plot,
       width = 11, height = 7, dpi = 300)

all_years_plot
```

## Average rental price in 2023
```{r}
zip_2023 = 
  nyc |>
  filter(year(date) == 2023) |>
  group_by(zip, borough) |>
  summarize(avg_2023 = mean(zori, na.rm = TRUE), .groups = "drop") |>
  mutate(avg_2023 = ifelse(is.nan(avg_2023), NA_real_, avg_2023)) |>
  drop_na(borough, avg_2023)

dist_2023_plot <- ggplot(zip_2023, aes(x = borough, y = avg_2023)) +
  geom_violin(trim = TRUE, alpha = 0.5) +
  geom_boxplot(width = 0.15, outlier.alpha = 0.3) +
  stat_summary(fun = median, geom = "point", size = 2) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Distribution of ZIP-level Average Rental Prices (2023)",
    subtitle = "Violin = distribution · Box = IQR · Dot = median",
    x = "Borough", y = "Average ZORI in 2023 ($)"
  ) +
  theme_minimal(base_size = 12)

ggsave("results/plot_2023_distribution.png", dist_2023_plot,
       width = 9, height = 6, dpi = 300)

dist_2023_plot

```
## Combine graph
```{r}
combined = 
  all_years_plot | dist_2023_plot +
  patchwork::plot_annotation(
    title = "NYC Rental Prices: Long-run Trends and 2023 Distributions"
  )

ggplot2::ggsave("results/combined_plots.png", combined,
                width = 17, height = 12, dpi = 300)
combined
```


# Problem 3

## Set-up data
```{r}
covar_path <- "data/nhanes_covar.csv"
accel_path <- "data/nhanes_accel.csv"

covar <- read_csv(covar_path) |> clean_names()
accel <- read_csv(accel_path) |> clean_names()
```

## Data preparation
```{r}
covar <- covar |> mutate(participant_id = row_number())
accel <- accel |> mutate(participant_id = row_number())

mims_cols <- names(accel) |> keep(~ str_detect(.x, "mims"))
accel <- accel |> mutate(across(all_of(mims_cols), as.numeric))

covar <- covar |>
  mutate(
    sex = case_when(
      x1 %in% c(1, "1") ~ "Male",
      x1 %in% c(2, "2") ~ "Female",
      TRUE ~ as.character(x1)
    ),
    age = as.numeric(x3),
    education = case_when(
      x4 %in% c(1, "1") ~ "Less than HS",
      x4 %in% c(2, "2") ~ "Some HS",
      x4 %in% c(3, "3") ~ "HS or GED",
      x4 %in% c(4, "4") ~ "Some college/AA",
      x4 %in% c(5, "5") ~ "College or more",
      TRUE ~ as.character(x4)
    ),
    sex = factor(sex, levels = c("Male","Female")),
    education = fct_relevel(
      factor(education),
      "Less than HS","Some HS","HS or GED","Some college/AA","College or more"
    )
  )

covar_keep <- covar |> filter(!is.na(age), age >= 21, !is.na(sex), !is.na(education))
merged <- covar_keep |> inner_join(accel, by = "participant_id")
```

## Table: number of men and women in each education category
```{r}
counts_tbl <- merged |>
  count(education, sex, name = "n") |>
  arrange(education, sex)
kable(counts_tbl, caption = "Number of Men and Women by Education")
```

## Plot on age distributions
```{r}
ggplot(merged, aes(x = age, fill = sex)) +
  geom_histogram(binwidth = 5, alpha = 0.6, position = "identity") +
  facet_wrap(~ education, ncol = 3, scales = "free_y") +
  labs(title = "Age distributions by sex within education levels",
       x = "Age (years)", y = "Count", fill = "Sex")
```

## total activity vs age
```{r}
merged <- merged |> mutate(total_activity = rowSums(across(all_of(mims_cols)), na.rm = TRUE))

ggplot(merged, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE, method = "loess") +
  facet_wrap(~ education, ncol = 3, scales = "free_y") +
  labs(title = "Total daily activity vs age by education",
       x = "Age (years)", y = "Total activity (sum of 24h MIMS)",
       color = "Sex")
```



## 24-hour activity time course
```{r}
tidy_long <- merged |>
  select(participant_id, sex, age, education, all_of(mims_cols)) |>
  pivot_longer(all_of(mims_cols), names_to = "minute_var", values_to = "mims") |>
  mutate(minute = readr::parse_number(minute_var))

minute_curves <- tidy_long |>
  group_by(education, sex, minute) |>
  summarize(mean_mims = mean(mims, na.rm = TRUE), .groups = "drop")

ggplot(minute_curves, aes(x = minute, y = mean_mims, color = sex)) +
  geom_line() +
  facet_wrap(~ education, nrow = 1, scales = "free_y") +
  labs(title = "24-hour activity time courses by education",
       x = "Minute of day (0–1440)", y = "Mean MIMS", color = "Sex")
```




