---
title: "HW3"
author: "wd2311"
date: "2025-10-14"
output: github_document
---
# Loading libraries
```{r}
library(tidyverse)
library(janitor)
library(p8105.datasets)
library(knitr)
library(forcats)
library(patchwork)
library(scales)
```


# Problem 1
## Set up and brief description
The instacart dataset contains 1,384,617 observations and 15 variables, each 
representing a single product added to a user’s grocery order on the Instacart 
online shopping platform. Each row corresponds to one product within one order, 
identified by variables such as:
order_id (unique order identifier),
product_id and product_name (the specific item purchased),
add_to_cart_order (the position in which the item was added),
reordered (whether the product had been purchased before by the same user),
order_dow and order_hour_of_day (day of week and time the order was placed),
aisle and department (product category classification).

```{r}
data("instacart") 

instacart |>
  summarize(
    n_obs = n(),
    n_vars = ncol(across()),
    var_names = paste(names(across()), collapse = ", ")
  )

glimpse(instacart)
```


## Description on aisles
There are 134 aisles. The aisles with the greatest number of items ordered are 
“fresh vegetables” (150609 orders) and “fresh fruits” (150473 orders).
```{r}
aisle_counts = 
  instacart |>
  count(aisle, sort = TRUE)

n_aisles = nrow(aisle_counts)
n_aisles

# top aisles
aisle_counts |> 
  slice_head(n = 10) |> 
  kable()
```

## Plot on aisles with more than 10000 items
```{r}
aisle_counts |>
  filter(n > 10000) |>
  mutate(aisle = fct_reorder(aisle, n)) |>
  ggplot(aes(x = n, y = aisle)) +
  geom_col() +
  labs(
    x = "Number of items ordered",
    y = NULL,
    title = "Items ordered by aisle (aisles with > 10,000 items)"
  ) +
  theme_minimal(base_size = 12)
```

## Table for top3 items by aisle
```{r}
top3_by_aisle = 
  instacart |>
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  count(aisle, product_name, sort = TRUE) |>
  group_by(aisle) |>
  slice_max(n, n = 3, with_ties = FALSE) |>
  arrange(aisle, desc(n)) |>
  ungroup()

kable(top3_by_aisle, col.names = c("Aisle", "Product", "Times ordered"))
```

## Table for mean hours of the day
```{r}
dow_labs = c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")

mean_hour_tbl = 
  instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  mutate(dow = factor(order_dow, levels = 0:6, labels = dow_labs)) |>
  group_by(product_name, dow) |>
  summarise(mean_hour = mean(order_hour_of_day), .groups = "drop") |>
  mutate(mean_hour = round(mean_hour, 2)) |>
  pivot_wider(names_from = dow, values_from = mean_hour) |>
  arrange(product_name)

kable(mean_hour_tbl, col.names = c("Product","Sun","Mon","Tue","Wed","Thu","Fri","Sat"))

```


# Problem 2
## Data Set-up
```{r}
zip_meta = 
  read_csv("data/Zip Codes.csv") |> 
  clean_names() 

zori_raw = 
  read_csv("data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv", show_col_types = FALSE) |> 
  clean_names() 

# Check column names
names(zip_meta)
names(zori_raw)
```

## Data preparation 
```{r}
zip_meta = 
  zip_meta |>
  mutate(
    zip = str_pad(as.character(zip_code), 5, pad = "0"),
    county = str_to_title(county) |> str_trim()
  )

county_to_borough <- c(
  "Bronx"          = "Bronx",
  "Kings"          = "Brooklyn",
  "New York"       = "Manhattan",
  "Queens"         = "Queens",
  "Richmond"       = "Staten Island"
)

zip_meta = 
  zip_meta |>
  mutate(borough = recode(county, !!!county_to_borough)) |>
  filter(!is.na(borough)) |>
  select(zip, borough, county, neighborhood)
```

## Turning to long form
```{r}
zip_lookup =
  zip_meta |>
  filter(!is.na(zip), !is.na(borough)) |>
  distinct(zip, borough) |>
  group_by(zip) |>
  slice(1) |>          
  ungroup()

zori_long = 
  zori_raw |>
  mutate(
    zip = str_pad(as.character(region_name), 5, pad = "0")
  ) |>
  pivot_longer(
    cols = matches("^x\\d{4}_\\d{2}_\\d{2}$"),
    names_to = "date_raw",
    values_to = "zori"
  ) |>
  mutate(
    date  = ymd(str_replace_all(str_remove(date_raw, "^x"), "_", "-")),
    year  = year(date),
    month = month(date)
  ) |>
  filter(!is.na(date))

nyc = 
  zori_long |>
  left_join(zip_lookup, by = "zip") |>
  filter(!is.na(borough))
```



## Monthly Observation
ZIP codes observed exactly 116 times: 149 
ZIP codes observed fewer than 10 times: 0 
```{r}
nyc_target = 
  nyc |>
  filter(date >= ymd("2015-01-01"), date <= ymd("2024-08-31"))

months_per_zip = 
  nyc_target |>
  count(zip, name = "n_months")

n_116  = months_per_zip |> 
  filter(n_months == 116) |> 
  nrow()
n_lt10 = months_per_zip |> 
  filter(n_months < 10)   |> 
  nrow()

cat("ZIP codes observed exactly 116 times:", n_116, "\n")
cat("ZIP codes observed fewer than 10 times:", n_lt10, "\n")

```

## Table on Borough × Year 
Average rents rise from 2015 to 2024 with a brief drop in 2020/2021.
Manhattan demonstrates the most expensive rents and Bronx the lowest.
```{r}
dir.create("results", showWarnings = FALSE)

borough_year_tbl = 
  nyc_target |>
  group_by(borough, year) |>
  summarize(avg_rent = mean(zori, na.rm = TRUE), .groups = "drop") |>
  arrange(borough, year)

borough_year_wide = 
  borough_year_tbl |>
  pivot_wider(names_from = year, values_from = avg_rent) |>
  mutate(across(-borough, ~ round(.x, 0))) 

kable(
  borough_year_wide |>
    mutate(across(-borough, ~ dollar(.x))),
  caption = "Average Rental Price by Borough and Year (ZORI)"
)


```


## Plot for all available years
Rents generally rise from 2015 to 2024, with a little drop in the year of 2021.
Manhattan is consistently the highest with most smooth curve, demonstrate low 
short-time variation.
```{r}
nyc_clean = 
  nyc_target |>
  filter(!is.na(zori), is.finite(zori))  

all_years_plot = 
  nyc_clean |>
  group_by(borough, date) |>
  mutate(borough_median = median(zori, na.rm = TRUE)) |>
  ungroup() |>
  ggplot(aes(x = date, y = zori, group = zip)) +
  geom_line(alpha = 0.08) +
  stat_summary(
    aes(group = borough),
    fun = median, geom = "line", linewidth = 1.1
  ) +
  facet_wrap(~ borough, ncol = 2, scales = "free_y") +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "NYC Rental Prices within ZIP Codes (All Years)",
    subtitle = "Thin lines: ZIP series · Thick line: Borough median ZORI",
    x = NULL, y = "ZORI ($)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())

ggsave("results/plot_all_years_by_borough.png", all_years_plot,
       width = 11, height = 7, dpi = 300)

all_years_plot
```

## Average rental price in 2023
In 2023, ZIP-level average rents are highest in Manhattan followed by Brooklyn. 
Manhattan demonstrated widest spread and a right-skewed upper tail, indicating 
substantial within-borough variation and high-end ZIPs.Other areas also shows a 
generally right skewed figure.Staten Island demonstrates lowest variation.
```{r}
zip_2023 = 
  nyc |>
  filter(year(date) == 2023) |>
  group_by(zip, borough) |>
  summarize(avg_2023 = mean(zori, na.rm = TRUE), .groups = "drop") |>
  mutate(avg_2023 = ifelse(is.nan(avg_2023), NA_real_, avg_2023)) |>
  drop_na(borough, avg_2023)

dist_2023_plot <- ggplot(zip_2023, aes(x = borough, y = avg_2023)) +
  geom_violin(trim = TRUE, alpha = 0.5) +
  geom_boxplot(width = 0.15, outlier.alpha = 0.3) +
  stat_summary(fun = median, geom = "point", size = 2) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Distribution of ZIP-level Average Rental Prices (2023)",
    subtitle = "Violin = distribution · Box = IQR · Dot = median",
    x = "Borough", y = "Average ZORI in 2023 ($)"
  ) +
  theme_minimal(base_size = 12)

ggsave("results/plot_2023_distribution.png", dist_2023_plot,
       width = 9, height = 6, dpi = 300)

dist_2023_plot

```
## Combine graph
```{r}
combined = 
  all_years_plot | dist_2023_plot +
  patchwork::plot_annotation(
    title = "NYC Rental Prices: Long-run Trends and 2023 Distributions"
  )

ggsave("results/combined_plots.png", combined,
                width = 17, height = 12, dpi = 300)
combined
```


# Problem 3
## Read-in
```{r}
covar_raw = 
  read_csv("data/nhanes_covar.csv", skip = 4, show_col_types = FALSE) |>
  clean_names()
accel_raw = 
  read_csv("data/nhanes_accel.csv",  show_col_types = FALSE) |> 
  clean_names()
```

## Tidy demongraphic data
```{r}
covar = 
  covar_raw |>
  clean_names() |>
  mutate(
    seqn = suppressWarnings(as.numeric(seqn)),
    age  = suppressWarnings(as.numeric(age)),
    bmi  = suppressWarnings(as.numeric(bmi)),
    sex = case_when(
      sex == 1 ~ "Male",
      sex == 2 ~ "Female",
      str_to_lower(as.character(sex)) %in% c("m","male") ~ "Male",
      str_to_lower(as.character(sex)) %in% c("f","female") ~ "Female",
      TRUE ~ NA_character_
    ),
    education = case_when(
      education == 1 ~ "Less than HS",
      education == 2 ~ "High school",
      education == 3 ~ "More than HS",
      str_detect(str_to_lower(as.character(education)), "less") ~ "Less than HS",
      str_detect(str_to_lower(as.character(education)), "high") ~ "High school",
      str_detect(str_to_lower(as.character(education)), "more") ~ "More than HS",
      TRUE ~ NA_character_
    )
  ) |>
  filter(
    !is.na(seqn), !is.na(sex), !is.na(age), !is.na(bmi), !is.na(education),
    age >= 21
  ) |>
  mutate(
    sex = factor(sex, levels = c("Male","Female")),
    education = factor(education,
                       levels = c("Less than HS","High school","More than HS"),
                       ordered = TRUE)
  ) |>
  distinct(seqn, .keep_all = TRUE)
```

## Tidy accelerometer data
```{r}
accel_core = 
  accel_raw |>
  clean_names() |>
  mutate(seqn = suppressWarnings(as.numeric(seqn))) |>
  distinct(seqn, .keep_all = TRUE)

minute_cols = 
  names(accel_core)[grepl("^min\\d{1,4}$", names(accel_core))]

accel_core = 
  accel_core |>
  mutate(
    across(
      all_of(minute_cols),
      ~ suppressWarnings(as.numeric(.x))
    )
  ) |>
  mutate(
    across(
      all_of(minute_cols),
      ~ ifelse(!is.na(.x) & .x < 0, NA_real_, .x)
    )
  )

accel_long = 
  accel_core |>
  pivot_longer(
    cols = all_of(minute_cols),
    names_to  = "minute_var",
    values_to = "mims"
  ) |>
  mutate(
    minute = suppressWarnings(as.integer(stringr::str_remove(minute_var, "^min")))
  ) |>
  filter(!is.na(minute), minute >= 1, minute <= 1440) |>
  mutate(
    time_of_day = as.POSIXct("2000-01-01 00:00:00", tz = "UTC") + lubridate::dminutes(minute - 1L)
  ) |>
  select(seqn, minute, time_of_day, mims)

```

## Merge
```{r}
nhanes_long = 
  covar |>
  inner_join(accel_long, by = "seqn")
```


## Table on Counts of Men and Women by Education
The table shows a balanced gender distribution across education levels. 
Men slightly outnumber women in the High School group (60.3% vs. 39.7%), 
while other groups are nearly equal.
```{r}
sex_by_edu = 
  covar |>
  count(education, sex, name = "Count") |>
  group_by(education) |>
  mutate(Percent = Count / sum(Count)) |>
  ungroup() |>
  mutate(Percent = percent(Percent, accuracy = 0.1)) |>
  arrange(education, sex)

kable(
  sex_by_edu,
  caption = "Counts of Men and Women by Education (NHANES accelerometer sample, ≥21y)",
  align = c("l", "l", "r", "r")
)
```



## Visualization of age distributions
The density plots indicate similar age distributions between men and women 
within each education level. Most participants are between ages 50–70, with the 
More than High School group showing a slightly younger range. Overall, no major 
age differences are observed between sexes.
```{r}
p_age_density = 
  ggplot(covar, aes(x = age, fill = sex, color = sex)) +
  geom_density(alpha = 0.25) +
  facet_wrap(~ education, nrow = 1, scales = "free_y") +
  labs(
    title = "Age distributions by sex within education levels",
    x = "Age (years)", y = "Density", fill = "Sex", color = "Sex"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

p_age_density
```

## Total Daily activities vs. Age
Across all education levels, total daily activity tends to decline with age. 
Men and women show similar overall patterns, though women appear slightly more 
active in the High School and More than High School groups. The wide spread of 
points indicates large individual variation beyond age and sex differences.
```{r}
totals = 
  nhanes_long %>%
  group_by(seqn) %>%
  summarize(total_activity = sum(mims, na.rm = TRUE), .groups = "drop") %>%
  inner_join(covar, by = "seqn")  

p_total_vs_age = 
  ggplot(totals, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.6, size = 1.6) +
  geom_smooth(se = FALSE, method = "loess", span = 0.9) +
  facet_wrap(~ education, nrow = 1, scales = "free_y") +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  labs(
    title = "Total daily activity vs. age",
    subtitle = "Colored by sex; faceted by education level",
    x = "Age (years)",
    y = "Total daily activity (sum of minute-level MIMS)",
    color = "Sex"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

p_total_vs_age
```
## 24 hour activity time courses by education
```{r}
diurnal_mean = 
  nhanes_long %>%
  group_by(education, sex, minute) %>%
  summarize(mean_mims = mean(mims, na.rm = TRUE), .groups = "drop") %>%
  mutate(time_of_day = as.POSIXct("2000-01-01 00:00:00", tz = "UTC") + dminutes(minute - 1L))

p_diurnal = 
  ggplot(diurnal_mean, aes(x = time_of_day, y = mean_mims, color = sex)) +
  geom_line(alpha = 0.7, linewidth = 0.6) +
  geom_smooth(se = FALSE, method = "loess", span = 0.1, linewidth = 0.9) +
  facet_wrap(~ education, nrow = 1, scales = "free_y") +
  scale_x_datetime(
    labels = date_format("%H:%M"),
    breaks = date_breaks("4 hours"),
    minor_breaks = date_breaks("1 hour")
  ) +
  labs(
    title = "24-hour activity time courses by education",
    subtitle = "Mean minute-level MIMS; color indicates sex",
    x = "Time of day", y = "Mean MIMS", color = "Sex"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

p_diurnal
```
Across all education groups, activity levels rise sharply in the morning, stay high throughout daytime, and drop at night. This consistent 24-hour rhythm reflects typical waking and sleeping patterns. Women show slightly higher mean activity during daytime—especially in the High school and More than HS groups—while men display a steeper evening decline in the Less than HS group. Overall, diurnal patterns are similar across sexes and education levels.

