---
title: "HW3"
author: "wd2311"
date: "2025-10-14"
output: github_document
---
# Loading libraries
```{r}
library(tidyverse)
library(janitor)
library(p8105.datasets)
library(knitr)
library(forcats)
```


# Problem 1
## Set up and brief description
The instacart dataset contains 1,384,617 observations and 15 variables, each 
representing a single product added to a user’s grocery order on the Instacart 
online shopping platform. Each row corresponds to one product within one order, 
identified by variables such as:
order_id (unique order identifier),
product_id and product_name (the specific item purchased),
add_to_cart_order (the position in which the item was added),
reordered (whether the product had been purchased before by the same user),
order_dow and order_hour_of_day (day of week and time the order was placed),
aisle and department (product category classification).
```{r}
data("instacart") 
glimpse(instacart)
n_obs  <- nrow(instacart)
n_vars <- ncol(instacart)
n_aisles <- instacart |> 
  distinct(aisle_id, aisle) |> 
  nrow()

c(n_obs, n_vars, n_aisles)
```

## description on aisles
There are 134 aisles. The aisles with the greatest number of items ordered are 
“fresh vegetables” (150,609 orders) and “fresh fruits” (150,473 orders).
```{r}
aisle_counts <- instacart |>
  count(aisle, sort = TRUE)

n_aisles <- nrow(aisle_counts)
n_aisles

# top aisles
aisle_counts |> slice_head(n = 10) |> kable()
```
## Plot on aisles with more than 10000 items
```{r}
aisle_counts |>
  filter(n > 10000) |>
  mutate(aisle = fct_reorder(aisle, n)) |>
  ggplot(aes(x = n, y = aisle)) +
  geom_col() +
  labs(
    x = "Number of items ordered",
    y = NULL,
    title = "Items ordered by aisle (aisles with > 10,000 items)"
  ) +
  theme_minimal(base_size = 12)
```

## Table for top3 items by aisle
```{r}
top3_by_aisle <- instacart |>
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  count(aisle, product_name, sort = TRUE) |>
  group_by(aisle) |>
  slice_max(n, n = 3, with_ties = FALSE) |>
  arrange(aisle, desc(n)) |>
  ungroup()

kable(top3_by_aisle, col.names = c("Aisle", "Product", "Times ordered"))
```

## Table for mean hours of the day
```{r}
dow_labs <- c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")

mean_hour_tbl <- instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  mutate(dow = factor(order_dow, levels = 0:6, labels = dow_labs)) |>
  group_by(product_name, dow) |>
  summarise(mean_hour = mean(order_hour_of_day), .groups = "drop") |>
  mutate(mean_hour = round(mean_hour, 2)) |>
  pivot_wider(names_from = dow, values_from = mean_hour) |>
  arrange(product_name)

kable(mean_hour_tbl, col.names = c("Product","Sun","Mon","Tue","Wed","Thu","Fri","Sat"))

```






```{r}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```

